---
title: "tSNE"
output: html_document
---

This is set to run automatically first in a new session, even If I ran another code-chunk.
```{r setup, include=FALSE}
# show results as well as errors of chunks in export
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(tidymodels)
library(Rtsne)
```

## 1. t-SNE plot with the iris dataset

### 1.1 Let's check an example

I use the `iris` dataset for that, which ships with the tidyverse package.

```{r}
# remove duplicates
iris_unique <- unique(iris)

# select variable columns into new object
iris_matrix <- as.matrix(iris_unique[,1:4])

# set seed for reproducibility
set.seed(42)

# calculate tsne values
tsne_iris_results <- Rtsne(iris_matrix,
                      pca = FALSE,
                      perplexity = 30,
                      theta = 0.0)

# plot foboga_results
plot(
  # take x and Y values from the  tsne_iris_results list Y data
  tsne_iris_results$Y,
  # color those points by species from original dataframe
  col = iris_unique$Species,
  # even unit lengths for both axis 
  asp = 1
  )
```

### 1.2 Let's try `ggplot`

#### 1.2.1 So it does not run right away

"`tsne_iris_results` is not a dataframe and cant be coerced by `fortify`."

```{r}
tsne_iris_results |> 
  ggplot(aes(x = Y, y = Y, colour = iris_unique$Species)) +
  geom_point()
```

#### 1.2.2 What does `fortify()` do?

```{r}
fortify(tsne_iris_results)
```

#### 1.2.3 cant be coerced `as.data.frame()` either

```{r}
as.data.frame(tsne_iris_results)
```

#### 1.2.4 A `list` seems to be something special within R.

Parts of the list can be used as a dataframe, by addressing these with `$`

```{r}
as.data.frame(tsne_iris_results$Y)
```

#### 1.2.5 Good, so try this with `ggplot()`

```{r}
ggplot(
  as.data.frame(tsne_iris_results$Y),
  aes(x = V1, y = V2, color = iris_unique$Species)
  ) +
  geom_point()
```

#### 1.2.6 Clean code a bit

```{r}
as.data.frame(tsne_iris_results$Y) |> 
  ggplot(aes(V1, V2, color = iris_unique$Species)) +
  geom_point()
```

## 2 Foboga data

Import the `foboga.csv` as `foboga` object.
Use either the `Import Dataset` -> `from Text (readr)` on top of the `Environment`-Tab, or `File`-> `Import Dataset` -> `from Text (readr)`.

Set columns `Timestamp` as `character` and `Station` as `numeric` respectively.

### 2.1 Basic EDA

#### 2.1.1 Scatterplot and Stats

```{r}
# from dataset foboga, plot Soil moisture over Humidity and color points by station
foboga |>
  # change column station from numeric to factor (works better for categorical data)
  mutate(Station = as.factor(Station)) |> 
  ggplot(aes(`Humidity (2m)`, `Soil moisture (-0.25m)`)) +
  # set point's opacity to 75%% to see more from the overlapping points
  geom_point(aes(colour = Station, alpha = 0.25)) +
  # set a more approachable color palette
  scale_color_brewer(palette = "Set1")

# from dataset foboga for each station,
# calculate number of entries, mean humidity and mean soil moisture
foboga |> 
  group_by(Station) |> 
  summarise(n = n(),
            mean_humidity2m = mean(`Humidity (2m)`),
            mean_soil_moisture = mean(`Soil moisture (-0.25m)`))
```

### 2.2 Tidy data

#### 2.2.1 Convert columns `Timestamp` and `Station`

Time comes as `chr` data and is more useful as `num`. Station is `num` and is more useful as `factor`.

```{r}
foboga_tidy <- foboga |>
    mutate(Timestamp = ymd_hms(Timestamp),
           Station = as.factor(Station)
           )
```

#### 2.2.1 Humidity-Curve over time per Station

```{r}
# from dataset foboga_tidy
foboga_tidy |> 
  # plot "Humidity (2m)" over Timestamp and color by "Station"
  ggplot(aes(Timestamp, `Humidity (2m)`, color = Station)) +
  # fit an plot an linear model, show confidence interval 0.95
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_brewer(palette = "Set1")
```

### 2.3 What does Rtsne do?

#### 2.3.1 Select and Filter

Rtsne wants only unique/distinc rows and I also want the `Station` column to not be considered, as this is the category for later exploration.

```{r}
foboga_data <- foboga_tidy |>
  # filter unique entries
  distinct() |>
  # select all columns from Timestamp till Soil moisture
  select(`Timestamp` : `Soil moisture (-0.55m)`)

foboga_station <- foboga_tidy |> 
  distinct() |> 
  select(Station)
```

#### 2.3.2 Calculate Rtsne data

```{r}
## set seed for reproducibility
set.seed(42)

## calculate and safe rtsne data as objects
## Perplexity 50
foboga_Rtsne_perp50_the50 <- Rtsne(foboga_data,
                      perplexity = 50,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)
## Perplexity 100
foboga_Rtsne_perp100_the50 <- Rtsne(foboga_data,
                      perplexity = 100,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)
## Perplexity 200
foboga_Rtsne_perp200_the50 <- Rtsne(foboga_data,
                      perplexity = 200,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)
## Perplexity 1000
foboga_Rtsne_perp1000_the50 <- Rtsne(foboga_data,
                      perplexity = 1000,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)

```

Yeah, This could be solved with a function. But I haven't gone down that road, yet.

#### 2.3.3 Plot

```{r}
## Perplexity 50
as.data.frame(foboga_Rtsne_perp50_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")

## Perplexity 100
as.data.frame(foboga_Rtsne_perp100_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")

## Perplexity 200
as.data.frame(foboga_Rtsne_perp200_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")

## Perplexity 1000
as.data.frame(foboga_Rtsne_perp1000_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")
```

#### 2.3.4 Tidy some more

So with the `Timestamp` consideres we can only see, that there are different clusters over time. So the timestamp explains the most variability. Loose it and see how the other variables interact better.

```{r}
foboga_select <- foboga_data |> 
  select(`Temperature (2m)` : `Soil moisture (-0.55m)`)

```

#### 2.3.5 Calculate some more

```{r}
## calculate and safe Rtsne data as objects
## Perplexity 50
foboga_select_Rtsne_perp50_the50 <- Rtsne(foboga_select,
                      perplexity = 50,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)
## Perplexity 100
foboga_select_Rtsne_perp100_the50 <- Rtsne(foboga_select,
                      perplexity = 100,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)
## Perplexity 200
foboga_select_Rtsne_perp200_the50 <- Rtsne(foboga_select,
                      perplexity = 200,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)
## Perplexity 1000
foboga_select_Rtsne_perp1000_the50 <- Rtsne(foboga_select,
                      perplexity = 1000,
                      theta = 0.5,
                      check_duplicates = FALSE,
                      num_threads = 0)
```

#### 2.3.6 Plot some more

```{r}
## Perplexity 50
as.data.frame(foboga_select_Rtsne_perp50_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")

## Perplexity 100
as.data.frame(foboga_select_Rtsne_perp100_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")

## Perplexity 200
as.data.frame(foboga_select_Rtsne_perp200_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")

## Perplexity 1000
as.data.frame(foboga_select_Rtsne_perp1000_the50$Y) |> 
  ggplot(aes(V1, V2, color = as.factor(foboga_station$Station)))+
  geom_point(aes(alpha = 0.25)) +
  scale_color_brewer(palette = "Set1")
```

### 2.4 PCA

[R Studio PCA Tutorial ](https://www.statology.org/principal-components-analysis-in-r/)

#### 2.4.1 Calculate Principal Components

```{r}
# calculate principal components
foboga_results <- prcomp(foboga_select, scale. = TRUE)

# reverse the signs
## don't know why 
foboga_results$rotation <- -1*foboga_results$rotation

# display composition of principal components
foboga_results$rotation

```

#### 2.4.2 Scores

```{r}
# reverse the signs of the scores
## still dont know why
foboga_results$x <- -1*foboga_results$x

# check scores for the first few rows
head(foboga_results$x)
```

#### 2.4.3 Plot

[Biplot with GGplot](https://www.geeksforgeeks.org/how-to-create-a-biplot-in-r/)

```{r}
# ggfortify can handle PCA-Data
library(ggfortify)

autoplot(foboga_results,
         data = foboga_select,
         color = (foboga_station$Station),
         loadings = TRUE,
         loadings.label = TRUE,
         loadings.label.size = 3,
         alpha = 0.25) +
  scale_color_brewer(palette = "Set1")
```
